skater: An R package for SNP-based Kinship Analysis, Testing, and Evaluation

### 2021-09-07 paper/paper.Rmd
remove references to supplementary material

---
title: "`skater`: An R package for SNP-based Kinship Analysis, Testing, and Evaluation"

@@

```{r, echo=FALSE, eval=FALSE}
## Run this chunk interactively if you update the package vignette.
## This renders the vignette as a PDF with a new title.

@@

The skater package has several functions for importing, parsing, and analyzing pedigree data. Pedigrees define familial relationships in a hierarchical structure. Many genomics tools for working with pedigrees start with a .fam file, which is a tabular format with one row per individual and columns for unique IDs of the mother, father, and the family unit. The skater package contains the function `read_fam()` to read in a PLINK-formatted .fam file and another function `fam2ped()` to convert the content into a pedigree object as a nested tibble with one row per family. All pedigree processing from skater internally leverages a data structure from the kinship2 package [@sinnwell2014]. Further functions such as `plot_pedigree()` produce a multi-page PDF drawing a diagram of the pedigree for each family, while `ped2kinpair()` produces a pairwise list of relationships between all individuals in the data with the expected kinship coefficients for each pair (see Supplementary Material).

@@

Once estimated kinship is converted to degree, it may be of interest to compare the inferred degree to known degrees of relatedness. When aggregated over many relationships and inferences, this can help benchmark performance of a particular kinship analysis method. The skater package adapts a `confusion_matrix()` function from @clark2021 to provide standard contingency table metrics (e.g. sensitivity, specificity, PPV, precision, recall, F1, etc.) with a new reciprocal RMSE (R-RMSE) metric. The R-RMSE metric is defined more thoroughly in the skater package vignette (see Supplementary Material) and may be a preferable measure of classification accuracy when benchmarking relationship degree estimation. In many kinship benchmarking analyses, classification error is treated in a categorical manner (exact match plus or minus one degree), neglecting the true amount of sharing as a real number. Taking the reciprocal of the target and predicted degree in a typical RMSE calculation results in larger penalties for more egregious misclassifications (e.g., classifying a first-degree relative pair as second-degree) than misclassifications at more distant relationships (e.g., classifying a fourth-degree relative pair as fifth-degree).

@@

The skater R package provides a robust software package for data import, manipulation, and analysis tasks typically encountered when working with SNP-based kinship analysis tools. All package functions are internally documented with examples, and the package contains a vignette demonstrating usage, inputs, outputs, and interpretation of all key functions (see Supplementary Material). The package contains internal tests that are automatically run with continuous integration via GitHub Actions whenever the package code is updated. The skater package is permissively licensed (MIT) and is easily extensible to accommodate outputs from new genome-wide relatedness and IBD segment methods as they become available.

>>>>

---
title: "skater: An R package for SNP-based Kinship Analysis, Testing, and Evaluation"

@@

```{r, echo=FALSE, eval=FALSE}
## Run this chunk interactively if you update the package vignette as supplementary material.
## This renders the vignette as a PDF with a new title.

@@

The skater package has several functions for importing, parsing, and analyzing pedigree data. Pedigrees define familial relationships in a hierarchical structure. Many genomics tools for working with pedigrees start with a .fam file, which is a tabular format with one row per individual and columns for unique IDs of the mother, father, and the family unit. The skater package contains the function `read_fam()` to read in a PLINK-formatted .fam file and another function `fam2ped()` to convert the content into a pedigree object as a nested tibble with one row per family. All pedigree processing from skater internally leverages a data structure from the kinship2 package [@sinnwell2014]. Further functions such as `plot_pedigree()` produce a multi-page PDF drawing a diagram of the pedigree for each family, while `ped2kinpair()` produces a pairwise list of relationships between all individuals in the data with the expected kinship coefficients for each pair (see skater package vignette).

@@

Once estimated kinship is converted to degree, it may be of interest to compare the inferred degree to known degrees of relatedness. When aggregated over many relationships and inferences, this can help benchmark performance of a particular kinship analysis method. The skater package adapts a `confusion_matrix()` function from @clark2021 to provide standard contingency table metrics (e.g. sensitivity, specificity, PPV, precision, recall, F1, etc.) with a new reciprocal RMSE (R-RMSE) metric. The R-RMSE metric is defined more thoroughly in the skater package vignette and may be a preferable measure of classification accuracy when benchmarking relationship degree estimation. In many kinship benchmarking analyses, classification error is treated in a categorical manner (exact match plus or minus one degree), neglecting the true amount of sharing as a real number. Taking the reciprocal of the target and predicted degree in a typical RMSE calculation results in larger penalties for more egregious misclassifications (e.g., classifying a first-degree relative pair as second-degree) than misclassifications at more distant relationships (e.g., classifying a fourth-degree relative pair as fifth-degree).

@@

The skater R package provides a robust software package for data import, manipulation, and analysis tasks typically encountered when working with SNP-based kinship analysis tools. All package functions are internally documented with examples, and the package contains a vignette demonstrating usage, inputs, outputs, and interpretation of all key functions. The package contains internal tests that are automatically run with continuous integration via GitHub Actions whenever the package code is updated. The skater package is permissively licensed (MIT) and is easily extensible to accommodate outputs from new genome-wide relatedness and IBD segment methods as they become available.
